\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2016-11-09 doc/publications/plos16/plos16.tex
changed to US spelling

\section{Introduction}
\marginpar{
We have parallelize and parallelise.
US spelling. Rob
}

The digital revolution is seeing a dramatic increase in data collected about every aspect of life~\cite{Loebbecke2015}.
These data sets are not only growing vertically, by capturing more events, but also horizontally by capturing
more information about these events.  The challenge of big and ``wide'' data is especially pronounced in the health
space where, for example, whole genome sequence (WGS) technology enables researchers to interrogate all 3 billion base pairs of the
human genome.

@@

A tree based model is capable of taking this interaction of variables
into account. In addition, random forests are well suited for
processing ``wide'' genomic data for two reasons.  Firstly, while other machine learning applications have the
propensity to overfit datasets with more features $p$ than samples $n$ (a consequence of the ``curse of
dimensionality''~\cite{Bauer2014, bellman1961adaptive}), decision trees are resistant to overfitting. 
Secondly, random forests are also very easy to parallelise. As the forest is a sum of decision trees, it it possible to grow
separate trees on different processors and combine the trees. See the
discussion of the properties of random forests in Section \ref{section:methods}.

The most popular random forest software packages are R-based, although current implementations do not scale well with
increasing feature size~\cite{Wright.and.Ziegle.2016}.  This is despite R now supporting long vectors (length greater
than $2^{31}-1$), and the {\sc randomForest} implementation in R lending itself to parallelization to grow multiple trees
simultaneously~\cite{Liaw.and.Weiner.2002} and them combine them.  More suitable implementations for ``wide'' genomic data have been
developed in other programming languages. {\sc CloudForest}~\cite{Bressler2015} written in Go, achieves fast run
times by effective use of the CPU cache, optimising for different classes of features and efficiently multi-threading.
Similarly, {\sc Ranger}~\cite{Wright.and.Ziegle.2016} written in C++ with an R front-end is specifically optimised for
dealing with large data sets.

However, the use of traditional compute infrastructure limits the parallelization strategies that can be employed by
these methods.  The programs are limited to utilising only CPUs that are on the same computer node (multi-threading) or
else they run on CPUs distributed across nodes by virtue of the fact that there is no communication between the
processes (a separate tree is grown on each node).  Hadoop/Spark overcomes these limitation by enabling programs to
scale beyond compute-node boundaries and hence enable more sophisticated parallelization strategies.  In the case of
random forests this means that the computations for each node of a tree can be handed off to separate processors.

Here we introduce \cursedforest, a Hadoop/Spark based implementation of random forests specifically designed to cater
for ``big'' (many samples) and ``wide'' (many features) data sets. A Spark application runs on a ``driver'' node and
distributes tasks to the many ``worker'' nodes, or ``executors''. Using this facility, \cursedforest\ is capable of
parallelising the split for each node in a tree thereby handling millions of features, such as in the 1000 Genomes
Project data, which consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012}.  By also
utilising Spark to read in and manipulate the standard genomic variant format (VCF) directly, \cursedforest\ outperforms
existing tools even on small datasets where multi-threading generally performs well. Harnessing the virtually unlimited
capability to parallelise tasks, \cursedforest\ can build a large number of trees, which boosts the accuracy for
variable importance selection~\cite{Szymczak2016}.

We previously demonstrated the versatility and scalability of Spark by developing VariantSpark~\cite{OBrien2015}, a
framework allowing users to easily analyse Variant Call Format (VCF) files using ML algorithms on the Spark framework.
Using VariantSpark, we successfully built a k-means model on the full $2500 \times 80$ million matrix to cluster
individuals by their ethnicity achieving an Adjusted Rand Index (ARI) of 0.84 (with 1 being perfect clustering and -1 random
clustering).

@@

\cursedforest\ is included in our previously developed framework for Spark-based genomic data analysis,
VariantSpark~\cite{OBrien2015}, which therefore now enables supervised and unsupervised machine learning. This offers a
comprehensive analysis toolkit that can scale for the future data demands, where the richness from WGS data can be
utilised to ``fill in'' or impute the information at previously unobserved genomic positions in older array
technology~\cite{Howie2012}.  This provides the potential to impute samples in the GWAs catalogue and generate datasets
of hundreds of thousands of individuals with millions of variants, highlighting the need for incorporating modern
compute paradigms to deal with these challenges.

@@

Biau {\it et al.}~\cite{Biau.2012} provide a proof that a random forest model will converge at a rate that depends on
the cardinality of the set $S$ of ``strong predictors'' rather than on the number of variables $p$. That is, given a
function $y=f(S)$ depending on a set of variables $S$, a random forest will converge to the true function $y$ with a
rate that depends on the size of $S$ and not the number of noise variables in the data set. If the size of $S$ is small
compared to the total number of variables, then the rate of convergence will be quicker. It is not clear what the size
of the set $S$ will be in a GWAs application. There are cases of small groups of SNPs with large effectsd and also of
conditions caused by the interaction of large groups of SNPs. 
\marginpar{check this. Denis}

@@

\begin{figure}[tbhp] 
 \begin{adjustwidth}{-1.00in}{0in}
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[totalheight=8cm]{./figs/variable_importance.png} 
    \caption{The sum of the variable importance of the 5 significant variables.} 
    \label{fig7:a} 
    \vspace{4ex}
  \end{subfigure}%% 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[totalheight=8cm]{./figs/out-of-bag-prediction-error.png}
    \caption{OOB classification error} 
    \label{fig7:b} 
    \vspace{4ex}
  \end{subfigure} 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[totalheight=8cm]{./figs/numnber_of_variables_found.png} 
    \caption{The number of the 5 significant variables found in the top 5.} 
    \label{fig7:a} 
    \vspace{4ex}
  \end{subfigure}%% 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[totalheight=8cm]{./figs/rbo.png}
    \caption{The  rank-biased-overlap measure.} 
    \label{fig7:b} 
    \vspace{4ex}
  \end{subfigure} 
  \caption{{\bf Variable importance (Mean Decrease Gini) and OOB (out of bag) classification error estimate for random
      forests with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}  Each measure is shown
    as a function of \mtry and ntree. {\bf a)} The sum of the varible importances for the 5 significant variables.  {\bf
      b)} OOB (out of bag) classification error. {\bf c)} the number of signifivant variables found in the top 5. 
     {\bf d)} The rank-biased-overlap \cite[]{Webber.et.al.2010}. }
  \label{figure:synth}
\end{adjustwidth}
\end{figure}

@@

Firstly we look at CurstedForest horizontal scalability for a medium size dataset of 2.5 million variables and 5000
samples, by varying \mtry\ fraction and the number of CPU cores allocated to the execution. The raw results are
presented in table \ref{table10} and visualized in figure \ref{figure:synth} below. Regardless of the number of cores used,
\cursedforest\ displays approximately linear dependency between the execution time and the \mtry\ parameter (presented
here as a fraction of the number of variables) [Fig 1 a).

>>>>

\section{Introduction}
The digital revolution is seeing a dramatic increase in data collected about every aspect of life~\cite{Loebbecke2015}.
These data sets are not only growing vertically, by capturing more events, but also horizontally by capturing
more information about these events.  The challenge of big and ``wide'' data is especially pronounced in the health
space where, for example, whole genome sequence (WGS) technology enables researchers to interrogate all 3 billion base pairs of the
human genome.

@@

A tree based model is capable of taking this interaction of variables
into account. In addition, random forests are well suited for
processing ``wide'' genomic data for two reasons.  Firstly, while other machine learning applications have the
propensity to overfit datasets with more features $p$ than samples $n$ (a consequence of the ``curse of
dimensionality''~\cite{Bauer2014, bellman1961adaptive}), decision trees are resistant to overfitting. 
Secondly, random forests are also very easy to parallelize. As the forest is a sum of decision trees, it it possible to grow
separate trees on different processors and combine the trees. See the
discussion of the properties of random forests in Section \ref{section:methods}.

The most popular random forest software packages are R-based, although current implementations do not scale well with
increasing feature size~\cite{Wright.and.Ziegle.2016}.  This is despite R now supporting long vectors (length greater
than $2^{31}-1$), and the {\sc randomForest} implementation in R lending itself to parallelization to grow multiple trees
simultaneously~\cite{Liaw.and.Weiner.2002} and them combine them.  More suitable implementations for ``wide'' genomic data have been
developed in other programming languages. {\sc CloudForest}~\cite{Bressler2015} written in Go, achieves fast run
times by effective use of the CPU cache, optimizing for different classes of features and efficiently multi-threading.
Similarly, {\sc Ranger}~\cite{Wright.and.Ziegle.2016} written in C++ with an R front-end is specifically optimized for
dealing with large data sets.

However, the use of traditional compute infrastructure limits the parallelization strategies that can be employed by
these methods.  The programs are limited to utilizing only CPUs that are on the same computer node (multi-threading) or
else they run on CPUs distributed across nodes by virtue of the fact that there is no communication between the
processes (a separate tree is grown on each node).  Hadoop/Spark overcomes these limitation by enabling programs to
scale beyond compute-node boundaries and hence enable more sophisticated parallelization strategies.  In the case of
random forests this means that the computations for each node of a tree can be handed off to separate processors.

Here we introduce \cursedforest, a Hadoop/Spark based implementation of random forests specifically designed to cater
for ``big'' (many samples) and ``wide'' (many features) data sets. A Spark application runs on a ``driver'' node and
distributes tasks to the many ``worker'' nodes, or ``executors''. Using this facility, \cursedforest\ is capable of
parallelization the split for each node in a tree thereby handling millions of features, such as in the 1000 Genomes
Project data, which consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012}.  By also
utilizing Spark to read in and manipulate the standard genomic variant format (VCF) directly, \cursedforest\ outperforms
existing tools even on small datasets where multi-threading generally performs well. Harnessing the virtually unlimited
capability to parallelize tasks, \cursedforest\ can build a large number of trees, which boosts the accuracy for
variable importance selection~\cite{Szymczak2016}.

We previously demonstrated the versatility and scalability of Spark by developing VariantSpark~\cite{OBrien2015}, a
framework allowing users to easily analyze Variant Call Format (VCF) files using ML algorithms on the Spark framework.
Using VariantSpark, we successfully built a k-means model on the full $2500 \times 80$ million matrix to cluster
individuals by their ethnicity achieving an Adjusted Rand Index (ARI) of 0.84 (with 1 being perfect clustering and -1 random
clustering).

@@

\cursedforest\ is included in our previously developed framework for Spark-based genomic data analysis,
VariantSpark~\cite{OBrien2015}, which therefore now enables supervised and unsupervised machine learning. This offers a
comprehensive analysis toolkit that can scale for the future data demands, where the richness from WGS data can be
utilized to ``fill in'' or impute the information at previously unobserved genomic positions in older array
technology~\cite{Howie2012}.  This provides the potential to impute samples in the GWAs catalogue and generate datasets
of hundreds of thousands of individuals with millions of variants, highlighting the need for incorporating modern
compute paradigms to deal with these challenges.

@@

Biau {\it et al.}~\cite{Biau.2012} provide a proof that a random forest model will converge at a rate that depends on
the cardinality of the set $S$ of ``strong predictors'' rather than on the number of variables $p$. That is, given a
function $y=f(S)$ depending on a set of variables $S$, a random forest will converge to the true function $y$ with a
rate that depends on the size of $S$ and not the number of noise variables in the data set. If the size of $S$ is small
compared to the total number of variables, then the rate of convergence will be quicker. It is not clear what the size
of the set $S$ will be in a GWAs application. There are cases of small groups of SNPs with large effects and also of
conditions caused by the interaction of large groups of SNPs. 
\marginpar{check this. Denis}

@@

\begin{figure}[tbhp] 
 \begin{adjustwidth}{-1.00in}{0in}
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[totalheight=8cm]{./figs/variable_importance.png} 
    \caption{The sum of the variable importance of the 5 significant variables.} 
    \label{fig7:a} 
    \vspace{4ex}
  \end{subfigure}%% 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[totalheight=8cm]{./figs/out-of-bag-prediction-error.png}
    \caption{OOB classification error} 
    \label{fig7:b} 
    \vspace{4ex}
  \end{subfigure} 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[totalheight=8cm]{./figs/numnber_of_variables_found.png} 
    \caption{The number of the 5 significant variables found in the top 5.} 
    \label{fig7:a} 
    \vspace{4ex}
  \end{subfigure}%% 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[totalheight=8cm]{./figs/rbo.png}
    \caption{The  rank-biased-overlap measure.} 
    \label{fig7:b} 
    \vspace{4ex}
  \end{subfigure} 
  \caption{{\bf Variable importance (Mean Decrease Gini) and OOB (out of bag) classification error estimate for random
      forests with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}  Each measure is shown
    as a function of \mtry and ntree. {\bf a)} The sum of the variable importances for the 5 significant variables.  {\bf
      b)} OOB (out of bag) classification error. {\bf c)} the number of significant variables found in the top 5. 
     {\bf d)} The rank-biased-overlap \cite[]{Webber.et.al.2010}. }
  \label{figure:synth}
\end{adjustwidth}
\end{figure}

@@

Firstly we look at CursedForest horizontal scalability for a medium size dataset of 2.5 million variables and 5000
samples, by varying \mtry\ fraction and the number of CPU cores allocated to the execution. The raw results are
presented in table \ref{table10} and visualized in figure \ref{figure:synth} below. Regardless of the number of cores used,
\cursedforest\ displays approximately linear dependency between the execution time and the \mtry\ parameter (presented
here as a fraction of the number of variables) [Fig 1 a).
